Forecasting Presidential Candidate Support during the Final Week before Election Day

### 2024-11-04 paper/paper.qmd
Completed model validation section and updated title and subtitle

---
title: "Forecasting the 2024 US Presidential Election"
author: 
  - Tianning He
  - Julia Lee
  - Shuangyuan Yang
thanks: "Code and data are available at: https://github.com/JuliaJLee/Forecasting_US_Election_2024.git"
date: today
date-format: long
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(tinytex)
library(knitr)
library(ggplot2)
library(dplyr)
library(arrow)
library(here)
```

@@

Putting these variables together, the analysis is able to construct a model that forecasts the percentage of support (pct) for Harris and Trump by considering high-quality pollsters, a population of voters who a more likely to vote on election day, polls from various states, poll sample sizes, and the times at which polls were conducted. An in-depth explanation as to why the variables above are selected is provided in (SECTION).

@@

Summary outputs for each model (one for Harris and another for Trump) along with model diagnostics to validate these models can be found in the Appendix (SECTION).

@@

With @fig-forecast, it appears as the presidential election will be a tight race as the forecasts of percentage of support for Harris and Trump in Week 13 are very similar. In the final week leading up to the election (October 27, 2024 to November 2, 2024), Harris is expected to receive more support while Trump is expected to receive less support in Week 13. The exact forecasts of the model are as follows:

@@

>>>>

---
title: "Forecasting Presidential Candidate Support during the Final Week before Election Day"
subtitle: "In the final week leading up to the 2024 US election, Kamala Harris is expected to lead the polls with 48.03% of support with Donald Trump following close behind with 47.98% of support."
author: 
  - Tianning He
  - Julia Lee
  - Shuangyuan Yang
thanks: "Code and data are available at: https://github.com/JuliaJLee/Forecasting_US_Election_2024.git"
date: today
date-format: long
format: pdf
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(tinytex)
library(knitr)
library(ggplot2)
library(dplyr)
library(arrow)
library(here)
library(modelsummary)
```

@@

Putting these variables together, the analysis is able to construct a model that forecasts the percentage of support (pct) for Harris and Trump by considering high-quality pollsters, a population of voters who a more likely to vote on election day, polls from various states, poll sample sizes, and the times at which polls were conducted. An in-depth explanation as to why the variables above are selected is provided in (@sec-model).

@@

Summary outputs for each model (one for Harris and another for Trump) along with model diagnostics to validate these models can be found in the Appendix (@sec-validation).

@@

@fig-forecast shows two additional points (one for Harris and the other for Trump), and these points represent the forecasted percentage of support for each candidate in Week 13. With this, it appears as though the presidential election will be a tight race as the forecasts of percentage of support for Harris and Trump in Week 13 are very similar. In the final week leading up to the election (October 27, 2024 to November 2, 2024), Harris is expected to receive more support while Trump is expected to receive less support in Week 13. The exact forecasts of the model are as follows:

@@

### Model for Harris

@tbl-harris shows the summary output of the linear model that predicts the percentage of support for Harris across weeks.
The low AIC, BIC, and RMSE values ensure that this model fits the data well. Here, the low R² value indicates that the time variable, "Week", does not explain much of the variability of pct for Harris.

```{r}
#| warning: false
#| echo: false
#| label: tbl-harris
#| tbl-cap: "Summary Table of the Regression Results for Harris"

model <- readRDS(here("models/harris_model.rds"))

modelsummary(harris_model)

Looking  at the Residuals vs. Predictor, Residuals vs. Fitted Values, and Normal Q-Q plots with @fig-residualspred, @fig-residualsfit, and @fig-normal, there are no systematic patters (e.g. fanning) and the residuals appear to follow a normal distribution with little deviation from the QQ-line in the Q-Q plot. This suggests that the assumptions of linearity, homoscedasticity, independence, and normality are satisfied by this model.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-residualspred
#| fig-cap: Showing the Residual vs. Predictor Plot of the Linear Model for Harris

## Residual Plots & Normal Q-Q Plot

### Find residuals

residuals <- harris_model$residuals

### Residuals vs. Predictor

plot(residuals ~ final_pooled_polls_harris$Week, main = "Residuals vs. Week", xlab = "Week", ylab = "Residuals")
```
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-residualsfit
#| fig-cap: Showing the Residual vs. Fitted Values Plot of the Linear Model for Harris

### Residuals vs. Fitted Values

fitted_values <- harris_model$fitted.values

plot(residuals ~ fitted_values, main = "Residuals vs. Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
```
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-normal
#| fig-cap: Showing the Normal Q-Q Plot of the Linear Model for Harris

### Normal Q-Q Plot

qqnorm(residuals)
qqline(residuals)
```

### Model for Trump

@tbl-trump shows the summary output of the linear model that predicts the percentage of support for Trump across weeks.
Similar to @tbl-harris, the model fits the data well and can provide more precise predictions as indicated by the low AIC, BIC, and RMSE values. The R² value is low for this model as well, implying that the time variable, "Week", does not explain much of the variability of pct for Trump.

```{r}
#| warning: false
#| echo: false
#| label: tbl-trump
#| tbl-cap: "Summary Table of the Regression Results for Trump"

model <- readRDS(here("models/trump_model.rds"))

modelsummary(trump_model)

Residual plots and a Normal Q-Q plot for this linear model are provided with @fig-trumprespred, @fig-trumpresfit, and @fig-trumpnormal. As there are no observable patterns within the residual plots (e.g. fanning) and the residuals closely follow the QQ-line in the Q-Q plot, this implies that the assumptions of linearity, homoscedasticity, independence, and normality are also satisfied by this model.

```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-trumprespred
#| fig-cap: Showing the Residuals vs. Predictor Plot of the Linear Model for Trump

## Residual Plots & Normal Q-Q Plot

### Find residuals

residuals <- trump_model$residuals

### Residuals vs. Predictor

plot(residuals ~ final_pooled_polls_trump$Week, main = "Residuals vs. Week", xlab = "Week", ylab = "Residuals")
```
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-trumpresfit
#| fig-cap: Showing the Residuals vs. Fitted Values Plot of the Linear Model for Trump

### Residuals vs. Fitted Values

fitted_values <- trump_model$fitted.values

plot(residuals ~ fitted_values, main = "Residuals vs. Fitted Values", xlab = "Fitted Values", ylab = "Residuals")
```
```{r}
#| echo: false
#| warning: false
#| message: false
#| label: fig-trumpnormal
#| fig-cap: Showing the Normal Q-Q Plot of the Linear Model for Trump

### Normal Q-Q Plot

qqnorm(residuals)
qqline(residuals)
```
